---
title: "Roots surface analysis"
format:
  html:
    css: css/styles.css  
    embed-resources: true
    theme: cosmo
    code-tools: true
    toc: true
    toc-depth: 4
execute: 
  warning: false
  message: false
  echo: false
---

To distinguish between mature and growing root zones, a temporal difference across 9-day intervals was applied

```{r}
#| echo: false
library(terra)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(purrr)
library(gratia)
library(mgcv)
if (!dir.exists("../output")) {
  dir.create("../output")
}
```

#### Data preparation

```{r}
df_pixel_count <- read.csv("../data/root_pixels_count.csv")
image_bank <- read.table("../output/image_bank.txt")


df <- df_pixel_count %>%
  filter(file %in% image_bank$V1)%>%
  mutate(image_name = sub("\\.png$", "", file)) %>%   
  separate(file, into = c("bloc", "position", "orientation","year","month", "day", "hour", "minute", "fileextension1"), 
           sep = "_|\\.|H|M.| |-", remove = FALSE)%>%
  mutate(scanner= as.factor(paste(position, orientation, sep = "_")),
         depth = as.factor(substr(scanner, 2, 3)),
         position = as.factor(substr(scanner, 1, 1)),
         orientation = as.factor(substr(scanner, 5, 5)))

duplicate_date <- df %>%
  group_by(scanner, year, month, day, hour) %>%
  reframe(has_00 = any(minute == "00")) %>% ungroup()

df <- df %>%
  left_join(duplicate_date, 
            by = c("scanner","year","month","day","hour")) %>%
  mutate(minute = if_else(minute == "01" & !has_00, "00", minute),
    date = ymd_hm(paste(year, month, day, hour, minute))
  ) %>%select(-has_00)
# Some timestamps are recorded at minute = 1 instead of minute = 0.In some cases both exist, in others only minute = 1 exists. We want to correct minute = 1 to 0 only if there is no existing row at minute = 0.

df <- df %>%
  filter(minute == "00", hour %in% c("00", "06", "12", "18"))%>%
  filter(date> "2024-03-01 00:00:00")
```

#### Conversion of surface areas (pixels \> cm²)

The images were acquired at a resolution of 1200 dpi, equivalent to **472.44 pixels/cm**. Root surfaces (initially in pixels) are converted to **cm²** using the following formula:

$$
Surface (cm^2) = Pixels / (472.44)^2
$$

```{r}
#| echo: true
dpi <- 1200
pixels_per_cm <- dpi / 2.54
df <- df %>%
  mutate(root_cm2_count = root_pixels_count / (pixels_per_cm^2))
```

#### Image bank viewing

The evolution of root surfaces over time is represented, separated according to the position of the scanner:

```{r}
ggplot(df, aes(date, root_cm2_count))+
  geom_point()+
  labs(x = "Date", y = "Root area (cm²)")+
  facet_wrap(~scanner, ncol = 4)
```

#### Quantification of apex surfaces according to time step

The differences in surface area between two successive images correspond to newly emerged portions of the roots, also known as apices. These young segments play a key role in the processes of absorption and exudation. A key question is therefore to determine the optimal time step for defining these growth areas in a meaningful way.

We examined the distribution of newly formed surfaces (surface delta at the apex) across time intervals ranging from 6 to 360 hours (15 days). Only positive surface increments were considered to represent effective growth. Negative deltas, which may reflect measurement variability or surface reduction, were set to zero and therefore do not contribute to the growth estimates :

```{r}
time_lags <- seq(6, 720, by = 6)

calculate_differences <- function(df, lags) {
  df <- df %>%
    arrange(date)
  
  map_dfr(lags, function(lag_hr) {
    df %>%
      mutate(lag_hours = lag_hr,
             date_target = date + hours(lag_hr)) %>%
      left_join(df %>%
                  select(date, root_cm2_count) %>%
                  rename(date_target = date,
                         root_cm2_count_lag = root_cm2_count),
                by = c("date_target")) %>%
      mutate(delta_cm2 = root_cm2_count_lag - root_cm2_count)
  })
}

df_diff <- df %>%
  group_by(bloc, position, depth, orientation)%>%
  group_modify(~calculate_differences(.x, time_lags))%>%
  ungroup()%>%
  filter(!is.na(delta_cm2))%>%
  mutate(delta_cm2 = if_else(delta_cm2 < 0, 0, delta_cm2))
```

The average apex areas as a function of time step are shown, along with their variability (standard deviation). An initial average is calculated by scanner, then the values are aggregated:

```{r}
plot1 <- df_diff %>%
  group_by(bloc, scanner, lag_hours) %>%
  mutate(mean_delta = mean(delta_cm2, na.rm = TRUE),
         sd_delta = sd(delta_cm2, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(lag_hours) %>%
  mutate(mean_delta = mean(mean_delta, na.rm = TRUE),
         sd_delta = sd(sd_delta, na.rm = TRUE)) %>%
  ggplot(aes(lag_hours, mean_delta)) +
  geom_vline(xintercept = 24 * 9, linetype = "dashed", color = "red") +
  geom_line() +
  labs(x = "Time step (hours)", y = "Average surface area of the apices (cm²)")

plot2 <- plot1 +
  geom_ribbon(aes(ymin = mean_delta - sd_delta,
                  ymax = mean_delta + sd_delta,
                  alpha = 0.1)) +
  geom_vline(xintercept = 24 * 9, linetype = "dashed", color = "red") +
  labs(y = "Average surface area ± SD (cm²)") +
  theme(legend.position = "none")

plot1
plot2
```

#### **Characterisation of growth dynamics**

**Assumption:** roots grow dynamically during a certain active phase, followed by a slowdown or stabilisation. This active growth phase results in an increase in the surface area occupied by the apices. Quantifying this surface area according to the time lag from which it is approached makes it possible to characterise the growth dynamics and identify the surface areas actually associated with this active growth phase.

**Objective:** to identify the time threshold at which growth slows down, by modelling the surface area occupied by newly emerged root portions according to the time lag from which they are calculated.

**Method:** Modelling of the apex surface as a function of time step using a generalised additive model (GAM). Calculation of the model derivative to identify inflection points. Detection of the first point where the derivative becomes negative (phase change \> transition between active growth and slowdown).

```{r}
mod <- gam(delta_cm2 ~ s(lag_hours),  random = list(scanner= ~1)
           ,data = df_diff)
deriv <- derivatives(mod, order=1)
draw(deriv)+
  geom_vline(xintercept = 24 * 9,
             linetype = "dashed",
             color = "red")
```
