---
title: "Questioning the influence of roots: A soil fauna perspective on rhizosphere space"
format:
  html:
    css: css/styles.css  
    embed-resources: true
    code-tools: true
    
execute: 
  warning: false
  message: false
  echo: false
---

## Setupt and data preparation

```{r}
# Load core libraries
library(tidyverse)
library(mgcv)
library(broom)
library(gratia)
library(lubridate)

# Global settings
set.seed(123)
if (!dir.exists("../output")) dir.create("../output")

# Ring surface function for density normalization
ring_surface <- function(r, width) {
  pi * ((r + width)^2 - r^2)}
```

```{r}
# --- DATA IMPORT & CLEANING ---
habitat_data_raw <- read.csv("../output/distance_data_cleaned.csv") %>%
  mutate(date = parse_date_time(date, orders = c("Y-m-d H:M:S", "Y-m-d"), tz = "UTC")) %>%
  filter(position != "B") # Remove control position if not needed

# --- TEMPORAL ALIGNMENT (Periods 1 to 7) ---
target_dates <- tibble(
  period = 1:7, 
  start_date = as.Date(c("2024-03-20", "2024-05-25", "2024-07-15", "2024-09-19", 
                         "2024-12-10", "2025-02-15", "2025-04-20"))
)

habitat_data_raw <- habitat_data_raw %>%
  crossing(target_dates) %>%                
  filter(date >= start_date) %>%            
  group_by(date) %>%                         
  slice_max(start_date, n = 1) %>%           
  ungroup() 

# --- SCALING & TRANSFORMATION ---
# Scale environmental predictors and abundance metrics
habitat_data_scaled <- habitat_data_raw %>%
  mutate(across(c(depth, period, position), as.factor)) %>%
  mutate(across(
    c(root_cm2_count, root_growth_cm2_count_24h, temp, hum, 
      temp_soil_amplitude, hum_soil_amplitude, invertebrate_count, predators_count),
    ~ as.numeric(scale(.x))
  ))

# Pivot to long format for Root-Type comparison (Growth vs Mature)
habitat_data_long <- habitat_data_scaled %>%
  select(
    image_name, temp, hum, period, depth, position, Predicted.class.levelup, root_cm2_count,
    growth_obs = distance_root_growth_obs,
    growth_exp = distance_root_growth_mean_exp,
    growth_sd  = distance_root_growth_sd_exp,
    mature_obs = distance_root_mature_obs,
    mature_exp = distance_root_mature_mean_exp,
    mature_sd  = distance_root_mature_sd_exp
  ) %>%
  pivot_longer(
    cols = starts_with(c("growth", "mature")),
    names_to = c("root_type", ".value"),
    names_sep = "_"
  ) %>%
  mutate(z_score = (obs - exp) / sd) %>% # Standardized Effect Size
  drop_na()
```

## 1. The rhizosphere effect : attraction and spatial signature 

### Global attraction index

Does fauna actively select the rhizosphere? We compare Observed vs. Expected distances.

```{r}
rhizo_results <- habitat_data_long %>%
  group_by(Predicted.class.levelup, position, root_type) %>%
  summarise(
    n = n(),
    med_obs = median(obs),
    med_exp = median(exp),
    avg_exp_sd = mean(sd, na.rm = TRUE),
    attr_strength = (med_exp - med_obs) / avg_exp_sd * 100,
    p_val = wilcox.test(obs, exp, exact = FALSE)$p.value,
    .groups = 'drop'
  ) %>%
  mutate(
    p_label = case_when(p_val < 0.001 ~ "***", p_val < 0.01 ~ "**", p_val < 0.05 ~ "*", TRUE ~ "ns"),
    is_sig = p_val < 0.05
  )

bg_zones <- tibble(xmin = c(0, -Inf),xmax = c(Inf, 0),ymin = -Inf,ymax = Inf, zone = c("Selection", "Avoidance"))

ggplot(rhizo_results, aes(x = attr_strength, y = Predicted.class.levelup)) +
  geom_rect(data = bg_zones,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = zone),
    inherit.aes = FALSE,
    alpha = 0.1) +
  geom_vline(xintercept = 0, color = "black", size = 1) +
  geom_point(aes(size = n, color = is_sig)) +
  geom_text(aes(label = p_label), vjust = -1, size = 3.5) +
  facet_grid(root_type ~ position) +
  scale_fill_manual(
    name = "Spatial pattern",
    values = c("Selection" = "#8fce00", "Avoidance" = "#cc0000")) +
  scale_color_manual(
    name = expression(italic(p) < 0.05),
    values = c("TRUE" = "black", "FALSE" = "grey60")) +
  labs(x = "Attraction Strength (%)",
    y = NULL,
    title = "Rhizosphere Attraction Index") +
  scale_x_continuous(
  expand = expansion(mult = 0.05, add = 2))+
  theme_minimal()+
  theme(
  panel.border = element_rect(color = "black", fill = NA, linewidth = 0.2),
  panel.grid.minor = element_blank(),
  #strip.background = element_rect(fill = "grey95", color = "black"),
  #legend.position = "bottom"
)

```

-   Positive Strength (\>0) + Significant: The root is a Hotspot. The fauna is actively selecting the rhizosphere.

-   Negative Strength (\<0) + Significant: The root is a Coldspot. This suggests avoidance (e.g., due to root exudate toxicity or predator presence).

-   Non-Significant ($p > 0.05$): The fauna is distributed randomly relative to the root; there is no "rhizosphere effect" for this group.

### Spatial signature (riz width)

We model animal density as a function of distance to identify the Rhizosphere Influence Zone (RIZ) using GAM derivatives.

```{r}
# SPATIAL BINNING PARAMETERS 
# We use a fine bin width (0.1 cm) to maintain high resolution. 
# /!\ Sensitivity to bin size is reduced later by using an offset in the GAM.
max_dist <- 10
bin_w <- 0.1 

# DATA PREPARATION & NORMALIZATION
# Calculate the number of animals per distance bin and the available area.
rhizo_decay_data <- habitat_data_long %>%
  # Filter out distances that are too large and biased by edge effects
  filter(obs <= max_dist) %>%
  # Create bins and assign observations to the midpoint of the bin
  mutate(dist_bin = floor(obs / bin_w) * bin_w + bin_w/2) %>%
  group_by(dist_bin, Predicted.class.levelup, position, root_type) %>%
  summarise(n_animals = n(), .groups = "drop") %>%
  # GEOMETRIC CORRECTION: Calculate the area of the concentric ring (annulus).
  # As distance (r) increases, the sampled area increases proportionally.
  # Area = pi * (R_outer^2 - R_inner^2)
  mutate(area = pi * ((dist_bin + bin_w/2)^2 - (dist_bin - bin_w/2)^2))

# ANALYSIS FUNCTION (GAM + RIZ EXTRACTION) 
extract_gam_stats <- function(df) {
  # Minimum data threshold to ensure model stability
  if (n_distinct(df$dist_bin) < 5 || sum(df$n_animals) < 10) return(NULL)
  
  # MODEL: Generalized Additive Model (GAM)
  # Family = quasipoisson
  # Offset = log(area): Treats the sampled area as an 'exposure' variable. 
  # This converts the raw count model into a density model (count/area) 
  # without the statistical pitfalls of using ratios directly.
  model <- tryCatch({
    gam(n_animals ~ s(dist_bin, k = 5), 
        offset = log(area), 
        data = df, 
        family = quasipoisson())
  }, error = function(e) return(NULL))
  
  if (is.null(model)) return(NULL)
  
  # RHIZOSPHERE INFLUENCE ZONE (RIZ) IDENTIFICATION 
  # We use the first derivative (slope) of the model to find where 
  # the influence of the root stops (the plateau).
  eval_pts <- data.frame(dist_bin = seq(0.05, 10, length.out = 200))
  derivs <- derivatives(model, data = eval_pts, type = "forward")
  
  # STATISTICAL TIPPING POINT:
  # The RIZ is defined as the distance where the 95% Confidence Interval 
  # of the derivative first includes zero. At this point, the density 
  # slope is no longer significantly different from a horizontal line.
  riz_val <- derivs %>%
    mutate(lower = .derivative - 1.96 * .se,
           upper = .derivative + 1.96 * .se) %>%
    # Filter out the first 0.5cm to avoid noise near the root surface
    filter(dist_bin > 0.5, (lower < 0 & upper > 0) | lower > 0) %>% 
    slice(1) %>% 
    pull(dist_bin)
  
  # RESULTS SUMMARY 
  tibble(
    edf = summary(model)$edf,             # Complexity of the curve (1 = linear)
    p_val = summary(model)$s.pv,          # Significance of the spatial pattern
    dev_expl = summary(model)$dev.expl * 100, # % of variance explained by root distance
    riz_cm = if(length(riz_val) > 0) riz_val else NA_real_, # Distance of influence
    n_total = sum(df$n_animals)           # Total count for the group
  )
}

# EXECUTION 
# Group data by taxa and experimental position, then apply the function
rhizo_stats_table <- rhizo_decay_data %>%
  group_by(Predicted.class.levelup, position, root_type) %>%
  do(extract_gam_stats(.))

# Display the final summary table
knitr::kable(rhizo_stats_table)
```

**How to read the final table ?**

-   **Presence of a "Signature" (Statistical Significance):** The p-value of the smooth term $s(dist\_bin)$. If $p < 0.05$, there is a significant spatial structure. If $p > 0.05$, the animals are randomly distributed relative to the root (no signature).

-   **Complexity of the Attraction:** Effective Degrees of Freedom (EDF). $EDF \approx 1$: The attraction/repulsion is a simple linear gradient.$EDF > 1$ (e.g., 2 to 4): The relationship is non-linear (exponential decay or bimodal).1 This is the "classic" rhizosphere signature.$EDF > 4$: The pattern is highly complex (potentially reflecting micro-patches or noise).

-   **The "Extent" of Influence (The Zero-Crossing):** to define the Rhizosphere Width, find the distance where the first derivative of the GAM curve approaches zero (where the slope flattens). Look at your GAM plot; the distance where the confidence interval first overlaps with the "Bulk Soil" mean (the plateau) is your Rhizosphere Influence Zone (RIZ).\

## 2. The Geometric Niche: Habitat Availability Constraint

Does root saturation limit behavioral choice? As root density increases, Z-scores should converge to zero.

```{r}
# Model Z-score against root density
mod_z <- gam(z_score ~ s(root_cm2_count, k = 10) + Predicted.class.levelup + root_type, 
             data = habitat_data_long)

ggplot(habitat_data_long, aes(x = root_cm2_count, y = z_score)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -1.96, ymax = 1.96, fill = "grey80", alpha = 0.3) +
  geom_point(alpha = 0.1, aes(color = root_type)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), color = "black") +
  facet_grid(root_type ~ Predicted.class.levelup) +
  labs(x = "Root Density (scaled)", y = "Standardized Effect Size (Z-score)") +
  theme_minimal()

# Model output
z_summary <- habitat_data_long %>%
  group_by(Predicted.class.levelup, root_type) %>%
  summarise(
    n = n(),
    mean_z = mean(z_score),
    # Percentage of individuals actively selecting the rhizosphere
    prop_active_selectors = sum(z_score < -1.96) / n() * 100,
    # Correlation between density and choice intensity
    density_constraint_rho = cor(z_score, root_cm2_count, method = "spearman"),
    .groups = 'drop'
  )
z_summary
```

## 3. Environmental Drivers: Presence vs. Selection

We disentangle whether climate affects Abundance (occupancy) or Distance Residuals (active behavior).

```{r}
# 1. Aggregate data by image
env_analysis_data <- habitat_data_long %>%
  group_by(image_name, Predicted.class.levelup, temp, hum, root_cm2_count, position, period, depth, root_type) %>%
  summarise(abundance = n(), mean_z_score = mean(z_score), .groups = "drop")

# 2. Baseline: Remove effect of abundance/density on distance
baseline_mod <- gam(mean_z_score ~ s(abundance, k=5) + s(root_cm2_count, k=5), data = env_analysis_data)
env_analysis_data$dist_residuals <- residuals(baseline_mod)

# 3. Compare Importance: Abundance (Filter) vs. Residuals (Selection)
mod_abund_imp <- gam(abundance ~ s(temp) + s(hum) + s(root_cm2_count), 
                     data = env_analysis_data, family = nb())
mod_behav_imp <- gam(dist_residuals ~ s(temp) + s(hum) + s(root_cm2_count), 
                     data = env_analysis_data)

# Extract and plot relative importance
imp_comp <- bind_rows(
  tidy(mod_abund_imp) %>% mutate(model = "Presence (Abundance)"),
  tidy(mod_behav_imp) %>% mutate(model = "Selection (Residuals)")
)

ggplot(imp_comp, aes(x = term, y = statistic, fill = model)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Climate Importance: Presence vs. Selection", y = "F-statistic") +
  theme_minimal()
```
