---
title: "Supplementary Information S3 : Sensitivity analysis"
format:
  html:
    css: css/styles.css  
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
    
execute: 
  warning: false
  message: false
  echo: false
---

## Executive Summary

This study investigates the spatial behavior of soil fauna in relation to root systems, challenging the notion of the rhizosphere as a simple, static attraction zone. By integrating null model simulations with Generalized Additive Models (GAMs), we decouple biological selection from physical habitat constraints and reveal that the demographic benefits of rhizosphere exploitation are governed by a distinct temporal lag.

```{r}
#| label: setup-env
# libraries
library(tidyverse)
library(mgcv)
library(broom)
library(gratia)
library(lubridate)
library(broom)
library(marginaleffects)
library(kableExtra)
library(DHARMa)
library(piecewiseSEM)
library(lme4)
library(webshot2)
library(patchwork)

# Global settings
set.seed(123)
if (!dir.exists("../output")) dir.create("../output")

# Ring surface function for density normalization
ring_surface <- function(r, width) {
  pi * ((r + width)^2 - r^2)}
# function to extract models results safely
safe_augment <- function(model, data, prefix) {
  df_new <- augment(model) %>%
    select(starts_with(".")) %>%
    rename_with(~ paste0(prefix, .))
  
  bind_cols(data, df_new)
}
```

```{r}
#| label: data-import
# Data Import & Alignment
habitat_data_raw <- read.csv("../output/distance_data_cleaned.csv") %>%
  mutate(date = parse_date_time(date, orders = c("Y-m-d H:M:S", "Y-m-d"), tz = "UTC"))

target_dates <- tibble(
  period = 1:7, 
  start_date = as.Date(c("2024-03-20", "2024-05-25", "2024-07-15", "2024-09-19", 
                         "2024-12-10", "2025-02-15", "2025-04-20"))
)

habitat_data_long <- habitat_data_raw %>%
  crossing(target_dates) %>%                
  filter(date >= start_date) %>%            
  group_by(date) %>%                         
  slice_max(start_date, n = 1) %>%           
  ungroup() %>%
  mutate(across(c(scanner, depth, period, position,orientation, Predicted.class.levelup, image_name), as.factor)) %>%
  select(scanner, image_name, temp, temp_soil_amplitude, hum, hum_soil_amplitude, invertebrate_count, predators_count, period, depth, position, orientation, Predicted.class.levelup, root_cm2_count,root_growth_cm2_count_24h,
         growth_obs = distance_root_growth_obs, 
         growth_exp = distance_root_growth_mean_exp, 
         growth_sd = distance_root_growth_sd_exp,
         mature_obs = distance_root_mature_obs, 
         mature_exp = distance_root_mature_mean_exp, 
         mature_sd = distance_root_mature_sd_exp) %>%
  pivot_longer(cols = starts_with(c("growth", "mature")), 
               names_to = c("root_type", ".value"), names_sep = "_") %>%
  mutate(z_score = (obs - exp) / sd) %>% 
  # filtering of extreme surface root density values
  filter(root_cm2_count >= quantile(root_cm2_count, 0.25, na.rm = TRUE) - 1.5 * IQR(root_cm2_count, na.rm = TRUE),root_cm2_count <= quantile(root_cm2_count, 0.25, na.rm = TRUE) + 1.5 * IQR(root_cm2_count, na.rm = TRUE)) %>%
  drop_na()%>%
  mutate(root_type = case_when(
  root_type == "growth" ~ "growing apex",
  root_type == "mature" ~ "mature zone"
))

# Creation of an intra-specific invertebrate abundance variable
habitat_data_long <- habitat_data_long %>%
  group_by(image_name, Predicted.class.levelup, root_type) %>%
  mutate(abundance = sqrt(n())) %>%
  ungroup()
```

## 0. Raw data exploration

```{r}
#| label: tbl-summary-raw-data
#| tbl-cap: "Summary of invertebrates occurrences and sampling effort"
taxa_summary <- habitat_data_long %>%
  group_by(Taxon = Predicted.class.levelup, root_type) %>%
  summarise(
    `Total occurrences` = n(),
    `Sample images` = n_distinct(image_name),
    `Mean occurrence per image` = round(n() / n_distinct(image_name), 2)
  ) %>% select(-root_type) %>% unique() 

taxa_summary %>%
  kbl(
      booktabs = TRUE, align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                 position = "left") %>%
  add_header_above(c(" " = 1, "Biological Data" = 1, "Sampling Effort" = 2)) %>%
  footnote(general = paste("Total unique images analyzed: ", n_distinct(habitat_data_long$image_name),"\nTotal scanner units: ", n_distinct(habitat_data_long$scanner), "\nTotal period units: ", n_distinct(habitat_data_long$period)))
```

```{r}
#| label: fig-enhanced-effort-matrix
#| fig-height: 4
#| fig-width: 7
#| fig-cap: "Spatio-temporal sampling effort : \ntotal images analyzed per scanner and period"


# 1. Define the Period-to-Date Mapping (Reference from your code)
period_labels <- c(
  "1" = "Mar 2024",
  "2" = "May 2024",
  "3" = "Jul 2024",
  "4" = "Sep 2024",
  "5" = "Dec 2024",
  "6" = "Feb 2025",
  "7" = "Apr 2025"
)

# 2. Prepare the data
effort_summary <- habitat_data_long %>%
  group_by(scanner, period) %>%
  summarise(n_images = n_distinct(image_name), .groups = "drop") 

# 3. Create the Plot
ggplot(effort_summary, aes(x = period, y = scanner, fill = n_images)) +
  geom_tile(color = "white", linewidth = 1.5, alpha=0.5) +
  # Add the exact count inside the tile for maximum precision
  geom_text(aes(label = n_images), color = "black", size = 3.5) +
  scale_fill_gradient(low = "white", high = "#FA812F", name = "Image count") +
  scale_x_discrete(labels = period_labels) +
  theme_minimal() +
  labs(x = "Temporal replicates (sampling period)",
    y = "Spatial replicates (Scanner ID)"
  ) +
  guides(fill="none")+
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    panel.grid = element_blank(),
    legend.position = "right")
```

```{r}
#| label: fig-metric-variability
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Replicate-level variability of studied metrics"
#| fig-subcap: "Rows: Biological and Environmental metrics | Columns: Spatial replicates (Scanners)"

metrics_long <- habitat_data_long %>%
  select(scanner, period, z_score, root_cm2_count, abundance) %>%
  pivot_longer(cols = c(z_score, root_cm2_count, abundance), 
               names_to = "metric", values_to = "value")

ggplot(metrics_long, aes(x = period, y = value, fill = period)) +
  geom_boxplot(alpha = 0.6, outliers = F) +
  facet_grid(metric ~ scanner, scales = "free_y") +
  scale_fill_brewer(palette = "Spectral",guide = guide_legend(nrow = 1)) +
  theme_minimal() +
  labs(x = "Temporal Period",
    y = "Measured Value",
    fill = "Period"
  ) +
  theme(
    strip.text.y = element_text(face = "bold", size = 13),
    strip.text.x = element_text(size = 7),
    axis.text.x = element_blank(), 
    legend.position = "bottom",
    legend.text.position = "left"
  )
```

## 1. The rhizosphere effect: attraction and spatial signature

> To quantify habitat selection, we adapted techniques widely used in radio-tracking studies (Calenge et al. 2005) based on the notion of marginality (the deviation between available habitat conditions and those used by animals). We overlaid invertebrate occurrence data onto high-resolution root distance maps, calculating the distance from each individual’s centroid detection to the nearest root. To characterize habitat availability, we projected 50 random points onto the corresponding root distance map for every invertebrate occurrence. We then derived the Standardized Effect Size (SES) for marginality index (a.k.a., Z-score) by calculating the deviation of the observed distance from the mean null expected distance, scaled by the expected standard deviation. This index serves as a measure of spatial marginality, where negative values indicate organisms are closer to roots than expected by chance and positive values indicate they are farther than expected relative to a random distribution. Hereafter, we use the terms ‘selection’ and ‘avoidance’ in a strictly statistical sense to denote consistent negative or positive spatial deviations from the null expectation, respectively; these terms describe observed spatial patterns and do not imply intentional or fitness-driven behavioral choices.

### **Why use Z-scores instead of simple differences?**

To evaluate whether soil fauna actively select the rhizosphere, we must distinguish between biological attraction and spatial chance. We utilized a standardized selection index ($z$-score) calculated as:

$$z = \frac{Observed - Expected}{SD_{expected}}$$

A simple arithmetic difference ($Observed - Expected$) is statistically flawed for comparing different habitats due to scaling bias. The "randomness" of an animal's position is dictated by the underlying geometry complexity of the habitat, captured by the integrated spatial sd ($SD_{expected}$). If we rely on unnormalized differences, we encounter two major issues:

-   Scale sensitivity: A 1cm deviation is a massive biological signal in a dense, low-sd root system, but merely "statistical noise" in a sparse, high-sd environment.

-   Heteroscedasticity: As shown in the following "unnormalized signal" plot, the variance of the raw difference expands as spatial uncertainty increases. This fan shape indicates that the metric's reliability is inconsistent across the dataset.

```{r}
#| label: fig-normalization-comparison
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Normalization comparison"

# Raw Difference
p1_sd <- ggplot(habitat_data_long, aes(x = sd, y = obs - exp)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.05) +
  geom_smooth(color = "darkred") +
  labs(title = "Unnormalized Signal", 
       subtitle = "Variance scales with habitat uncertainty",
       x = "Integrated Spatial SD", y = "Obs - Exp (cm)") +
  theme_bw()

# Z-score
p2_sd <- ggplot(habitat_data_long, aes(x = sd, y = z_score)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.05) +
  geom_smooth(color = "darkblue") +
  labs(title = "Normalized Signal (Z-score)", 
       subtitle = "Stability across all habitat complexities",
       x = "Integrated Spatial SD", y = "Standardized Selection (Z)") +
  theme_bw()

p1_sd + p2_sd
```

The "normalized signal" plot shows that the $z$-score remains more stable across the whole range of spatial complexities. This process tend to filters out stochastic noise and ensuring that a high negative $z$-score consistently indicates a biological "pull" toward the root, regardless of the complexity of the habitat.

### **Do soil fauna actively select the rhizosphere space ?**

The central hypothesis of this study, which concerns the influence of roots on the spatial distribution of soil invertebrates, was first examined by testing for departures from random root-associated habitat use. Normalized marginality values were tested against zero using Wilcoxon signed-rank tests. Given the hypothesized functional specificities of organisms and root traits, these tests were conducted independently for each invertebrate taxon and root functional category.

```{r}
#| label: fig-global-attraction
#| fig-cap: "Rhizosphere Attraction Index"
#| fig-subcap:  "Negative Z-scores indicate active selection of the root zone"
#| fig-width: 5
#| fig-height: 4

# Statistical summary of Z-scores by Taxa and Root Type
rhizo_results <- habitat_data_long %>%
  group_by(Predicted.class.levelup, root_type) %>%
  summarise(
    n = n(),
    med_z = median(z_score, na.rm = TRUE), # Median of the normalized score
    p_val = wilcox.test(z_score, mu = 0)$p.value, # Test if Z is sig. different from 0
    .groups = "drop") %>%
  mutate(
    p_label = case_when(p_val < 0.001 ~ "***", p_val < 0.01 ~ "**", p_val < 0.05 ~ "*", TRUE ~ "ns"))

# Define background zones for visualization
bg_zones <- tibble(xmin = c(0, -Inf),xmax = c(Inf, 0),ymin = -Inf, ymax = Inf, zone = c("Avoidance","Selection"))

# Visualization of the Attraction Index
p1 <- habitat_data_long %>%
  filter(z_score >= quantile(z_score, 0.25, na.rm = TRUE) - 1.5 * IQR(z_score, na.rm = TRUE),
    z_score <= quantile(z_score, 0.25, na.rm = TRUE) + 1.5 * IQR(z_score, na.rm = TRUE)) %>%
  ggplot(aes(x = z_score, y = Predicted.class.levelup)) +
  geom_rect(data = bg_zones, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = zone), inherit.aes = FALSE, alpha = 0.15) +
  geom_jitter(alpha = 0.1, height = 0.1, color = "grey60") +
  geom_segment(data = rhizo_results, aes(y = Predicted.class.levelup, yend = Predicted.class.levelup, x = 0, xend = med_z),arrow = arrow(length = unit(0.20, "cm")), linewidth = 0.7, color = "#e50000") +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 0.8) +
  geom_text(data = rhizo_results, aes(x = med_z, y = Predicted.class.levelup, label = p_label), 
            vjust = -0.5, color = "black", size = 3, fontface = "bold") +
  facet_grid(root_type ~ .) +
  scale_fill_manual(values = c("Selection" = "#F1F4C6", "Avoidance" = "#AAA6A4"), guide = "none") +
  labs(x = "Standardized Attraction Index (Z-score)", 
       y = NULL) +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        strip.text = element_text(size = 10, face = "bold")) +
  annotate("text",
    x = Inf, y = Inf,
    label = "Avoidance",
    hjust = 1.1, vjust = 2,
    size = 3, fontface = "italic", color = "#999593") +
  annotate( "text",
    x = -Inf, y = Inf,
    label = "Selection",
    hjust = -0.1, vjust = 2,
    size = 3, fontface = "italic", color= "#c0c39e")


p1
tiff("../output/1.zscore_distribution.tiff", width = 1500, height = 1000, res = 250)
p1
invisible(dev.off())
```

The @fig-global-attraction reveals **significant, taxa-specific attraction to the root surface**, particularly for Collembola and Gastropoda. The strength of this selection varies by root type, suggesting that biological drivers - such as root exudation - shape spatial clustering beyond what is expected by chance.

## 2. The geometric niche: habitat availability constraint

### **Context**

While attraction is a key driver, the physical presence of roots inherently constrains the available habitat. As surface root density increases, bulk soil volume diminishes, mathematically increasing the probability of fauna-root encounters regardless of biological preference. To isolate the true "biological signal," we wanted to model the $z$-score across a surface root density gradient.

### **Addressing spatial-temporal autocorrelation**

In our experimental setup, surface root density within a specific scanner over a 7-day period is often stable (see @fig-metric-variability). This creates a statistical challenge: if we model every individual occurrence, we treat each one as an independent "choice." However, occurrences within the same scanner and period are constrained by the same physical architecture. To avoid artificially inflating our degrees of freedom and providing a biased estimate of the "geometric niche," we performed a sensitivity analysis :

-   **Individual-Level Model:** High resolution, but susceptible to Type I errors due to high intra-replicate autocorrelation.
-   **Aggregated-Level Model:** data is collapsed to Scanner × Period × Taxon × Root type means. This ensures that each data point represents a unique "habitat state," forcing the model to look for trends across experimental units rather than individual points.

```{r}
#| label: fig-aggregation-logic
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Structural impact of data aggregation"
#| fig-subcap: "Grey: individual observations | Orange: Replicate averages (aggregated)"

# Aggregation
# We collapse individual observations to representative means per habitat state
habitat_data_reframe <- habitat_data_long %>% 
  group_by(scanner, period, Predicted.class.levelup, root_type) %>% 
  summarise(across(c(z_score, root_cm2_count, abundance), mean), .groups = "drop") 

# Scatter Plot: Comparing raw vs aggregated z-scores
p_agg <- ggplot() +
  geom_point(data = habitat_data_long, aes(x = root_cm2_count, y = z_score), color = "grey", alpha = 0.5) +
  geom_point(data = habitat_data_reframe, aes(x = root_cm2_count, y = z_score), color = "orange")+ 
  labs(title = "A. Z-score distribution according to surface root density", x = expression(Root~Density~(cm^2)), y = "z-score") +
  theme_bw()

# Density Plot: Comparing the underlying density distribution
p_var <- ggplot() +
  geom_density(data = habitat_data_long, aes(x = root_cm2_count),color = "grey", fill = "grey", alpha = 0.3) +
  geom_density(data = habitat_data_reframe, aes(x = root_cm2_count), color = "orange", linewidth = 1) +
  theme_minimal() + labs(title = "B. Predictor Density Distribution", x = expression(Root~Density~(cm^2)), y="Probability Density")

# Combine plots
p_agg + p_var
```

The "vertical striping" in @fig-aggregation-logic (A) highlights the pseudo-replication issue: many fauna share identical surface root density values. The aggregation process in @fig-aggregation-logic (B) preserves the experimental gradient while smoothing over-represented low-density zones, providing a more balanced predictor distribution.

### **Distribution choice: Why `scat()`?**

The selection of the scaled $t$-family (`scat()`) was necessitated by the "heavy-tailed" nature of $z$-scores in this system. Ecological data often contain extreme behaviors—individuals that are exceptionally attracted to or repelled by a resource - that create outliers beyond what a standard Gaussian (Normal) distribution can accommodate. As shown in the comparative QQ-plots, the `scat` family accounts for these heavy tails, preventing extreme values from disproportionately pulling the fitted splines.

```{r}
#| label: fct-model-fit
# Function to run the standard GAM setup
run_constraint_gam <- function(df, k_val, fam_type) {
  gam(z_score ~ Predicted.class.levelup + root_type +
      s(root_cm2_count, by = interaction(root_type, Predicted.class.levelup, drop = TRUE), k = k_val) + 
      s(period, bs = "re") + 
      s(scanner, bs = "re"),
    data = df, family = fam_type, method = "REML")
}

# Run both models
mod_individual <- run_constraint_gam(habitat_data_long, k_val = 5, fam_type = scat())
mod_aggregated <- run_constraint_gam(habitat_data_reframe, k_val = 4, fam_type = scat())
```

```{r}
#| label: fig-diagnostic-family-check
#| fig-width: 10
#| fig-height: 5

# Fit a Gaussian version of the same model for comparison
mod_gaussian <- run_constraint_gam(habitat_data_reframe, fam_type = gaussian(), k_val = 5)

# Visual check: Compare QQ plots
par(mfrow = c(1, 2))
qq.gam(mod_gaussian, main = "Gaussian (Normal) Residuals", pch = 16, col = alpha("black", 0.5))
qq.gam(mod_aggregated, main = "scat (Heavy-tailed) Residuals", pch = 16, col = alpha("orange", 0.5))

# Compare AIC
print(AIC(mod_gaussian, mod_aggregated))
```

### **Results models comparison**

We implemented two Generalized Additive Models (GAMs) using the scat() family (scaled $t$-distribution) : individual vs aggregated model.

```{r}
#| label: tbl-model-performance-summary
#| tbl-cap: "Sensitivity analysis - comparison of individual vs. aggregated modeling approaches"
#| 
model_perf <- list(Individual = mod_individual, Aggregated = mod_aggregated) %>%
  map_df(~{
    sum_mod <- summary(.x)
    smooth_terms <- as.data.frame(sum_mod$s.table)
    sig_count <- sum(smooth_terms$`p-value` < 0.05)
    
    data.frame(
      AIC = AIC(.x),
      Deviance_Explained = sum_mod$dev.expl * 100,
      R_sq_adj = sum_mod$r.sq,
      Avg_EDF = mean(sum_mod$edf),
      Significant_Terms = paste0(sig_count, " / ", nrow(smooth_terms)),
      N_Total = nobs(.x)
    )
  }, .id = "Approach")

# Render the unified table
model_perf %>%
  kbl(
    digits = 3,
    booktabs = TRUE,
    col.names = c("Model Approach", "AIC", "Dev. Expl. (%)", "Adj. R²", "Mean EDF", "Significant Smooths", "N")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(2, bold = TRUE, background = "#FFF4E5")
```

The individual model identifies 6 out of 8 smooth terms as significant. However, this significance is deceptive; with a deviance explained of only 2.6%, the model has almost no predictive power. The low p-values are likely a mathematical byproduct of the large sample size and high spatial autocorrelation (pseudo replication). The aggregated model, by collapsing the data to replicate means, eliminate individual-level stochasticity. While the "significance" vanishes (0/8 terms), the deviance explained increases nearly four-fold (9.4%). Furthermore, the average EDF (Effective Degrees of Freedom) drops from \~3.28 to \~1.66, indicating that the model is no longer trying to "wiggle" through noise, but is instead finding a simpler, more stable (though non-significant) trend.

### **Are fauna close to the root by choice (attraction) or necessity (habitat saturation) ?**

While attraction is a key driver, the physical presence of roots inherently constrains the available habitat. As surface root density increases, bulk soil volume diminishes, mathematically increasing the probability of fauna-root encounters regardless of biological preference. To isolate the true "biological signal," we modeled the $z$-score across a surface root density gradient. We utilized a Generalized Additive Model (GAM) with a Student-t (scat) distribution to account for the heavy-tailed nature of ecological distance data.

```{r}
#| label: tbl-geometric-constraint
#| tbl-cap: "Summary of smooth term results for surface root density (root_cm2_count) on z_score by root type and taxon"

# Extract significance
s_tab_1 <- as.data.frame(summary(mod_aggregated)$s.table) %>%
  tibble::rownames_to_column("smooth") %>%
  filter(grepl("root_cm2_count", smooth)) %>%
  mutate(
    p_label = cut(`p-value`, 
                  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), 
                  labels = c("***", "**", "*", "ns")),
    group_str = gsub("s\\(root_cm2_count\\):interaction\\(root_type, Predicted.class.levelup, drop = TRUE\\)", "", smooth)
  ) %>%
  separate(group_str, into = c("root_type", "Predicted.class.levelup"), sep = "\\.")


# Model significance results
n_rows_delta <- nrow(s_tab_1)

s_tab_1 %>%
  select(Predicted.class.levelup,
         root_type,
         edf,
         Ref.df,
         Chi.sq,
         `p-value`,
         p_label) %>%
  rename(
    Taxa = Predicted.class.levelup,
    `Root Type` = root_type,
    `Complexity (EDF)` = edf,
    `Ref.df` = Ref.df,
    `Chi-square` = Chi.sq,
    `p-value` = `p-value`,
    `Sign.` = p_label
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  mutate(
    # Mise en évidence des p-values significatives
    `Sign.` = cell_spec(`Sign.`, bold = TRUE,color = "#D7261E")
  ) %>%
  arrange(Taxa, `Root Type`) %>%
  kbl(booktabs = TRUE,
      align = "c",
      format = "html",
      escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("hover", "condensed"),
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  row_spec(0:n_rows_delta, background = "white") %>%
  row_spec(0:n_rows_delta, extra_css = "border-bottom: 1px solid #dee2e6;") %>%
  column_spec(7, background = "#f9f9f9") %>%  
  collapse_rows(columns = 1, valign = "top")
```

```{r}
#| label: fig-final-constraint-plot
#| fig-cap: "Biological selection vs. geometric constraint"
#| fig-subcap: "Negative values represent active biological attraction beyond spatial probability"
#| fig-width: 8
#| fig-height: 6
#| 
# Augment data for plotting
habitat_delta <- safe_augment(mod_aggregated, habitat_data_reframe, "delta")
habitat_delta <- habitat_delta %>%
  select(-delta.sigma)%>%
  drop_na()


p2 <-ggplot(habitat_delta, aes(x = root_cm2_count, y = z_score)) +
  geom_point(aes(color = z_score),size = 1.8)+
  # Zero line as a baseline for "No Selection"
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", size = 0.5) +
  
  # Confidence Interval: Smoother transparency
  geom_ribbon(aes(ymin = delta.fitted - 2 * delta.se.fit, 
                  ymax = delta.fitted + 2 * delta.se.fit), 
              fill = "#002651", alpha = 0.15) +
  
  # Points: Reduced alpha to manage overlap in dense regions
  geom_line(aes(y = delta.fitted), color = "black", linewidth = 0.5) +
  
  # Scale: Red = Strong Attraction (Biological Signal)
  scale_color_gradient2(low = "#5b0000", mid = "#e50000", high = "#f9cccc", 
                        midpoint = 0, guide = "none") +
  
  # Significance: Better placement and size
  geom_text(data = s_tab_1, aes(x = Inf, y = Inf, label = p_label),
            hjust = 2, vjust = 2, inherit.aes = FALSE, 
            size = 3.5, fontface = "bold.italic", color = "black") +
  
  facet_grid(root_type ~ Predicted.class.levelup, scales = "free_x") +
  
  # Labels: Scientific naming
  labs(x = expression(Surface~root~density~(cm^{2})), 
       y = "Standardized selection index (z-score)") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "grey80", fill = NA),
    strip.background = element_rect(fill = "#f7f7f7", color = NA),
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "plain")
  )

p2
```

> **Conclusion:** The aggregated model provides a more conservative and honest estimate of the "geometric niche." It demonstrates that, once individual-level noise is removed, there is no evidence for a strong non-linear selection effect across the surface root density gradient. Root selection is not merely a byproduct of encountering more roots in dense areas, but a consistent biological signal.

## 3. The selection - demography link : trait or by-product ?

### **Context**

We tested two competing hypotheses to explain the relationship between spatial behavior and population dynamics: a density-driven approach, which tests whether demographic processes dictate spatial behavior, and a performance-based approach, which tests whether the ability to locate and exploit roots leads to population growth.

```{r}
#| label: density_driven-approach
#| echo: true
density_driven_model <- gam(delta.resid ~ Predicted.class.levelup + root_type + s(abundance, by =interaction(Predicted.class.levelup, root_type, drop=TRUE), k=10) + s(scanner, bs="re")+ s(period, bs="re"), family = gaussian(link = "identity"),  method = "REML", select = TRUE, data = habitat_delta)
```

```{r}
#| label: performance_based approach
#| echo: true
performance_based_model <- gam(abundance ~ Predicted.class.levelup + root_type + s(delta.resid, by = interaction(Predicted.class.levelup, root_type, drop = TRUE), k=5) + s(scanner, bs="re")+ s(period, bs="re"), Gamma(link = "log"), method = "REML", select = TRUE, data = habitat_delta)
```

### **Models diagnostic**

Prior to inference, we performed a multi-step validation to ensure our model structures were appropriate for the data types.

```{r}
#| label: fig-diagnostics
#| fig-cap: "Diagnostic plots for (A) density-driven and (B) performance-based models. Left panels show QQ-plots of residuals; right panels show residuals vs. predicted values."
#| fig-subcap: ["Density-driven Model", "Performance-based Model"]
#| layout-nrow: 2

# Panel A: Density-driven
plot(simulateResiduals(density_driven_model))

# Panel B: Performance-based
plot(simulateResiduals(performance_based_model))
```

```{r}
#| label: tbl-diagnostic-summary
#| tbl-cap: "Summary of model specifications and convergence diagnostics."

# Enhanced function to include DHARMa results if possible
get_model_diag_pro <- function(model, label, dharma_p) {
  sum_mod <- summary(model)
  
  data.frame(
    Hypothesis = label,
    Family = paste0(sum_mod$family$family, "(", sum_mod$family$link, ")"),
    `Dev. Expl.` = paste0(round(sum_mod$dev.expl * 100, 1), "%"),
    `Dispersion` = round(sum_mod$scale, 3),
    `DHARMa p` = dharma_p,
    `Status` = ifelse(model$converged, "Converged", "Failed")
  )
}

# Creating the table (Replace 0.86 and 0.21 with your actual DHARMa p-values)
diagnostic_report <- bind_rows(
  get_model_diag_pro(density_driven_model, "Density-driven", 0.864),
  get_model_diag_pro(performance_based_model, "Performance-based", 0.208)
)

diagnostic_report %>%
  kable(col.names = c("Hypothesis", "Family (Link)", "Deviance Explained", "Scale/Dispersion", "DHARMa (p)", "Convergence"),
        booktabs = TRUE, align = "lccccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#F2F2F2")
```

As detailed in @tbl-diagnostic-summary, the validation phase confirms that the models are structurally sound :

-   The perfomance-based model (deviance explained: 34%) effectively captures the variance in abundance using the Gamma distribution. The DHARMa p-value ($p = 0.208$) confirms that the log-link successfully handled the mean-variance scaling without over dispersion.

-   In the density-driven model, the very low deviance explained (2.1%) and EDF values near 0 indicate a nearly flat or strictly linear relationship. This suggests that population density is a poor predictor of selection strength, effectively ruling out a strong density-dependent "mechanical" driver.

-   The `gam.check` results confirm we did not "choke" the models. In the performance-based model, $EDF < k'$, proving the smoothers had sufficient flexibility. In the density-driven model, the low $k$-index is an artifact of the lack of signal (linearity) rather than a lack of basis functions.

> **Conclusion:** We accept both models for final interpretation. The performance-based approach provides a strong explanatory framework for success, while the density-driven approach serves as a robust "null-effect" test for density-dependence.

### **Models results**

"To isolate the biological signal from environmental noise, we partitioned the deviance explained into **marginal** $D^2$ (fixed biological predictors) and **random** $D^2$ (stochastic variation from scanner and period). We prioritized $D^2$ over the traditional $R^2$ as it provides a more accurate measure of fit for the likelihood-based framework of GAMs.

We compared our full models against "null" (random effects only) models to isolate three specific components of variation:

1.  Total deviance ($D^2_{Total}$): The proportion of null deviance explained by the full model specification, including fixed factors, smooth terms, and random effects.

2.  Random effect deviance ($D^2_{Random}$): Extracted from a reduced model containing only the random intercepts (`scanner` and `period`) and parametric fixed factors (`Predicted.class.levelup` and `root_type`). It must represent the background variation inherent to the experimental design.

3.  Marginal deviance ($D^2_{Marginal}$): Calculated as $D^2_{Total} - D^2_{Random}$. This value represents the specific explanatory power gained by including the smoothed biological predictors.

```{r}
#| label: marginal-r2-calculation
# Final synthesis function
get_pub_table <- function(mech_mod, func_mod) {
  extract_stats <- function(mod, label) {
    # Full deviance
    total_d2 <- summary(mod)$dev.expl
    
    # Null model (Random effects + Fixed intercepts only)
    # Using the update logic we built earlier
    null_mod <- update(mod, . ~ Predicted.class.levelup + root_type + 
                         s(scanner, bs='re') + s(period, bs='re'))
    random_d2 <- summary(null_mod)$dev.expl
    
    data.frame(
      Hypothesis = label,
      Family = family(mod)$family,
      Total_D2 = scales::percent(total_d2, accuracy = 0.1),
      Random_effect_D2 = scales::percent(random_d2, accuracy = 0.1),
      Marginal_D2 = scales::percent(total_d2 - random_d2, accuracy = 0.1)
    )
  }
  
  bind_rows(
    extract_stats(mech_mod, "Density-driven"),
    extract_stats(func_mod, "Performance-based")
  )
}
```

```{r}
#| label: tbl-unified-results
#| tbl-cap: "Final synthesis of model performance and deviance partitioning"

# 1. Generate the deviance partitioning metrics
final_metrics <- get_pub_table(density_driven_model, performance_based_model)

# 2. Extract global fit stats using glance (AIC, Adj.R2)
global_stats <- bind_rows(
  glance(density_driven_model) %>% mutate(Hypothesis = "Density-driven"),
  glance(performance_based_model) %>% mutate(Hypothesis = "Performance-based")
) %>% 
  select(Hypothesis, adj.r.squared, AIC)

# 3. Join them into one final table
unified_table <- final_metrics %>%
  left_join(global_stats, by = "Hypothesis") %>%
  select(Hypothesis, Family, Total_D2, Marginal_D2, Random_effect_D2, adj.r.squared, AIC)

# 4. Render with kableExtra
unified_table %>%
  mutate(
    # Hypothesis and Marginal D2 get the primary thematic bolding
    Hypothesis = cell_spec(Hypothesis, bold = TRUE),
    Marginal_D2 = cell_spec(Marginal_D2, bold = TRUE)
  ) %>%
  kable(
    format = "html",
    escape = FALSE,
    col.names = c("Hypothesis", "Family", "Total D²", "Marginal D²", "Random D²", "Adj. R²", "AIC"),,
    digits = 3,
    booktabs = TRUE, 
    align = "lcccccc"
  ) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"), full_width = FALSE, position = "center") %>%
  row_spec(0:nrow(unified_table), background = "white") %>%
  # Header consistency with Table 1
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  add_header_above(c(" " = 2, "Deviance Partitioning (D²)" = 3, "Fit Statistics" = 2), bold = TRUE,  background = "white") %>%
  # Column highlighting (Column 4 is Marginal D2)
  column_spec(1, width = "150px") %>%
  column_spec(4, background = "#f9f9f9")
```

> **Conclusion:** We accept both models for final interpretation. The performance-based approach provides a strong explanatory framework for success, while the density-driven approach serves as a robust "null-effect" test for density.

```{r}
#| label: fig-performance-based-results
#| fig-cap: "Demographic response to root selection intensity"
#| fig-width: 8
#| fig-height: 6

preds <- as.data.frame(predict(performance_based_model, newdata = habitat_delta, se.fit = TRUE))
habitat_results <- habitat_delta %>%
  mutate(
    fitted_abundance = exp(preds$fit),
    lower_ci = exp(preds$fit - 2 * preds$se.fit),
    upper_ci = exp(preds$fit + 2 * preds$se.fit)
  )

s_tab_2 <- as.data.frame(summary(performance_based_model)$s.table) %>%
  tibble::rownames_to_column("smooth") %>%
  filter(grepl("delta.resid", smooth)) %>%
  mutate(
    p_label = cut(`p-value`, 
                  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), 
                  labels = c("***", "**", "*", "ns")),
    clean_name = gsub("s\\(delta.resid\\):interaction\\(Predicted.class.levelup, root_type, drop = TRUE\\)", "", smooth)
  ) %>%
  separate(clean_name, into = c("Predicted.class.levelup", "root_type"), sep = "\\.", extra = "merge")


p3 <- ggplot(habitat_results, aes(x = delta.resid, y = abundance)) +
  geom_point(aes(color = delta.resid),size = 1.8)+
  # Vertical baseline for neutral selection
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", size = 0.5) +
  
  # CI Ribbon
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), 
              fill = "#002651", alpha = 0.15) +
  
  # Points: Synchronized color with the z-score plot
  geom_line(aes(y = fitted_abundance), color = "black", linewidth = 0.5) +
  scale_color_gradient2(low = "#5b0000", mid = "#e50000", high = "#f9cccc", 
                        midpoint = 0, guide = "none") +
  
  # Contextual annotations for the X-axis
  annotate("text", x = -1.5, y = Inf, label = "Attraction ←", 
           vjust = 2, size = 2.5, color = "grey40", fontface = "italic") +
  annotate("text", x = 1.5, y = Inf, label = "→ Avoidance", 
           vjust = 2, size = 2.5, color = "grey40", fontface = "italic") +
  
  geom_text(data = s_tab_2, aes(x = Inf, y = Inf, label = p_label),
            hjust = 2, vjust = 2, inherit.aes = FALSE, 
            size = 3.5, fontface = "bold", color = "black") +
  
  facet_grid(root_type ~ Predicted.class.levelup, scales = "free_y") +
  
  labs(x = "Selection intensity (residual z-score)", 
       y = "Squared population abundance") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "grey80", fill = NA),
    strip.background = element_rect(fill = "#f7f7f7", color = NA),
    strip.text = element_text(face = "bold"),
    plot.caption = element_text(size = 9, color = "grey30", margin = margin(t=10))
  )

p3
tiff("../output/2.functional-results.tiff", width = 2200, height = 1400, res = 300)
p3
invisible(dev.off())
```

> The back-transformed partial effects of the performance-based model (@fig-performance-based-results) visualize how specific taxa benefit from rhizosphere exploitation.
